---
title: "Homework 1"
date: "`r Sys.Date()`"
author: "Max Rohde"
output:
  rmdformats::robobook:
    highlight: kate
---

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(class)
```

## Q1
Paste the code from the mixture-data-lin-knn.R file into the homework template Knitr document.

```{r}
# load binary classification example data from author website 
## 'ElemStatLearn' package no longer available
load(url('https://web.stanford.edu/~hastie/ElemStatLearn/datasets/ESL.mixture.rda'))
dat <- ESL.mixture

plot_mix_data <- expression({
  plot(dat$x[,1], dat$x[,2],
       col=ifelse(dat$y==0, 'blue', 'orange'),
       pch=20,
       xlab=expression(x[1]),
       ylab=expression(x[2]))
  ## draw Bayes (True) classification boundary
  prob <- matrix(dat$prob, length(dat$px1), length(dat$px2))
  cont <- contourLines(dat$px1, dat$px2, prob, levels=0.5)
  rslt <- sapply(cont, lines, col='purple')
})

eval(plot_mix_data)

## fit linear classifier
fit_lc <- function(y, x) {
  x <- cbind(1, x)
  beta <- drop(solve(t(x)%*%x)%*%t(x)%*%y)
}

## make predictions from linear classifier
predict_lc <- function(x, beta) {
  cbind(1, x) %*% beta
}

## fit model to mixture data and make predictions
lc_beta <- fit_lc(dat$y, dat$x)
lc_pred <- predict_lc(dat$xnew, lc_beta)

## reshape predictions as a matrix
lc_pred <- matrix(lc_pred, length(dat$px1), length(dat$px2))
contour(lc_pred,
      xlab=expression(x[1]),
      ylab=expression(x[2]))


## find the contours in 2D space such that lc_pred == 0.5
lc_cont <- contourLines(dat$px1, dat$px2, lc_pred, levels=0.5)

## plot data and decision surface
eval(plot_mix_data)
sapply(lc_cont, lines)

## fit knn classifier
## use 5-NN to estimate probability of class assignment
knn_fit <- knn(train=dat$x, test=dat$xnew, cl=dat$y, k=5, prob=TRUE)
knn_pred <- attr(knn_fit, 'prob')
knn_pred <- ifelse(knn_fit == 1, knn_pred, 1-knn_pred)

## reshape predictions as a matrix
knn_pred <- matrix(knn_pred, length(dat$px1), length(dat$px2))
contour(knn_pred,
        xlab=expression(x[1]),
        ylab=expression(x[2]),
        levels=c(0.2, 0.5, 0.8))


## find the contours in 2D space such that knn_pred == 0.5
knn_cont <- contourLines(dat$px1, dat$px2, knn_pred, levels=0.5)

## plot data and decision surface
eval(plot_mix_data)
sapply(knn_cont, lines)


## do bootstrap to get a sense of variance in decision surface
resample <- function(dat) {
  idx <- sample(1:length(dat$y), replace = T)
  dat$y <- dat$y[idx]
  dat$x <- dat$x[idx,]
  return(dat)
}
  
## plot linear classifier for three bootstraps
par(mfrow=c(1,3))
for(b in 1:3) {
  datb <- resample(dat)
  ## fit model to mixture data and make predictions
  lc_beta <- fit_lc(datb$y, datb$x)
  lc_pred <- predict_lc(datb$xnew, lc_beta)
  
  ## reshape predictions as a matrix
  lc_pred <- matrix(lc_pred, length(datb$px1), length(datb$px2))

  ## find the contours in 2D space such that lc_pred == 0.5
  lc_cont <- contourLines(datb$px1, datb$px2, lc_pred, levels=0.5)
  
  ## plot data and decision surface
  eval(plot_mix_data)
  sapply(lc_cont, lines)
}

## plot 5-NN classifier for three bootstraps
par(mfrow=c(1,3))
for(b in 1:3) {
  datb <- resample(dat)
  
  knn_fit <- knn(train=datb$x, test=datb$xnew, cl=datb$y, k=5, prob=TRUE)
  knn_pred <- attr(knn_fit, 'prob')
  knn_pred <- ifelse(knn_fit == 1, knn_pred, 1-knn_pred)
  
  ## reshape predictions as a matrix
  knn_pred <- matrix(knn_pred, length(datb$px1), length(datb$px2))

  ## find the contours in 2D space such that knn_pred == 0.5
  knn_cont <- contourLines(datb$px1, datb$px2, knn_pred, levels=0.5)
  
  ## plot data and decision surface
  eval(plot_mix_data)
  sapply(knn_cont, lines)
}

## plot 20-NN classifier for three bootstraps
par(mfrow=c(1,3))
for(b in 1:3) {
  datb <- resample(dat)
  
  knn_fit <- knn(train=datb$x, test=datb$xnew, cl=datb$y, k=20, prob=TRUE)
  knn_pred <- attr(knn_fit, 'prob')
  knn_pred <- ifelse(knn_fit == 1, knn_pred, 1-knn_pred)
  
  ## reshape predictions as a matrix
  knn_pred <- matrix(knn_pred, length(datb$px1), length(datb$px2))
  
  ## find the contours in 2D space such that knn_pred == 0.5
  knn_cont <- contourLines(datb$px1, datb$px2, knn_pred, levels=0.5)
  
  ## plot data and decision surface
  eval(plot_mix_data)
  sapply(knn_cont, lines)
}
```


## Q2
Read the help file for R's built-in linear regression function `lm`

```{r}
?lm
```

## Q3
Re-write the functions `fit_lc` and `predict_lc` using `lm`, and the associated predict method for `lm` objects.

### New functions using `lm`
```{r}
## fit linear classifier
fit_lc <- function(y, x) {
  
  m <- cbind(y, x)
  colnames(m) <- c("y", "x1", "x2")
  
  df <- as_tibble(m)
  
  mod <- lm(y ~ x1 + x2, data=df)
  
  return(mod)
}

## make predictions from linear classifier
predict_lc <- function(mod, xnew) {
  colnames(xnew) <- c("x1", "x2")
  df_new <- as_tibble(xnew)
  predict(mod, df_new)
}
```

### Testing to see that code works with new functions
```{r}

## fit model to mixture data and make predictions
lc_mod <- fit_lc(dat$y, dat$x)
lc_pred <- predict_lc(lc_mod, dat$xnew)

## reshape predictions as a matrix
lc_pred <- matrix(lc_pred, length(dat$px1), length(dat$px2))
contour(lc_pred,
      xlab=expression(x[1]),
      ylab=expression(x[2]))


## find the contours in 2D space such that lc_pred == 0.5
lc_cont <- contourLines(dat$px1, dat$px2, lc_pred, levels=0.5)

## plot data and decision surface
eval(plot_mix_data)
sapply(lc_cont, lines)

```

## Q4
Consider making the linear classifier more flexible, by adding squared terms for x1 and x2 to the linear model

See modified functions with quadratic terms below.
```{r}
## fit linear classifier
fit_lc <- function(y, x) {
  
  m <- cbind(y, x)
  colnames(m) <- c("y", "x1", "x2")
  
  df <- as_tibble(m)
  
  mod <- lm(y ~ poly(x1,2) + poly(x2,2), data=df)
  
  return(mod)
}

## make predictions from linear classifier
predict_lc <- function(mod, xnew) {
  colnames(xnew) <- c("x1", "x2")
  df_new <- as_tibble(xnew)
  predict(mod, df_new)
}
```

Now we see that the contours are curved, rather than straight.
```{r}
## fit model to mixture data and make predictions
lc_mod <- fit_lc(dat$y, dat$x)
lc_pred <- predict_lc(lc_mod, dat$xnew)

## reshape predictions as a matrix
lc_pred <- matrix(lc_pred, length(dat$px1), length(dat$px2))
contour(lc_pred,
      xlab=expression(x[1]),
      ylab=expression(x[2]))


## find the contours in 2D space such that lc_pred == 0.5
lc_cont <- contourLines(dat$px1, dat$px2, lc_pred, levels=0.5)

## plot data and decision surface
eval(plot_mix_data)
sapply(lc_cont, lines)

```

## Q5
Describe how this more flexible model affects the bias-variance tradeoff

A more flexible model increases the variance but reduces the bias. This is because the model can fit more closely to the true classification contour, but this flexibility comes at the cost of increased variability over replications.

## Extra
I made an animation with ``gganimate`` to show the bias variance tradeoff with the LS and NN classifiers.

<video controls
    src="https://maximilianrohde.com/anim/knn_vs_ls.mp4"
    width="650">
</video>
